---
title: "Visualization of the best results of *SignalingProfiler* 2.0 benchmarking"
author: "Veronica Venafra"
output:
  html_document:
    theme: flat
    highlight: zenburn
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
```

## Introduction
This workbook refers to the manuscript *SignalingProfiler 2.0: a network-based approach to bridge multi-omic data to phenotypic hallmarks*.

## Prerequisites
To visualize the output of this chunks, open Cytoscape Desktop app. 

## Import libraries and functions

```{r message = FALSE}
source('0.libraries.R')
source('0.functions.R')

network_dir <- './results/4.network_generation/'
pheno_dir <- './results/5.phenoscore_benchmarking/'
```

## Import best results of Step2 and Step3

```{r}
# Read the best model object returned by SignalingProfiler
best_model_id <- 'Network1554'
best_object <- readRDS(paste0(network_dir, 'val_', best_model_id, '.RDS'))

# Define nodes table
nodes_df <- best_object$nodes_df

# Extract igraph object of best model
network_graph <- best_object$igraph_network
```

```{r}
best_model_id <- 'pheno89'
best_pheno_result <- readRDS(paste0(pheno_dir, 'best_', best_model_id, '.rds'))
```

## Visualize PhenoScore barplot

```{r fig.align='center', fig.width=5, fig.height=5}
best_pheno_result$barplot
```
**Figure 4B** Bar plot reporting the inferred modulation of phenotypes upon metformin treatment after Step3 of SignalingProfiler (activation in red, inhibition in blue).

## Cytoscape visualization

To visualize the model in Cytoscape using the different SignalingProfiler layout you have to properly format the attributes.
Moreover, we want to add as additional node attribute, whether a protein is a member of gold standard list.
```{r}
final_sp_visualization <- format_for_visualization(best_pheno_result$sp_object_phenotypes) 

regulators <- read_excel('./input/Gold standard.xlsx', sheet = 1, skip = 1) %>%
  mutate(Regulation = ifelse(grepl('MTOR*', Regulator), Regulation*-100, Regulation*100))

nodes_df_raw <- final_sp_visualization$nodes_df
nodes_df_mapped <- nodes_df_raw %>%
  mutate(gold_standard = ifelse(gene_name %in% regulators$Target, 'Y', 'N'))

# override formatted network
final_sp_visualization$igraph_network <- igraph::graph_from_data_frame(d = final_sp_visualization$edges_df, 
                                                                       vertices =  nodes_df_mapped)


#write_rds(final_sp_visualization, './results/5.phenoscore_benchmarking/formatted_output.rds')

```

### Visualize the whole network
This is the code used for **Figure S11** of the manuscript.
```{r eval = FALSE}
RCy3::createNetworkFromIgraph(igraph=final_sp_visualization$igraph_network, name="Whole network")

# Set in Cytoscape the SignalingProfiler style available in SignalingProfiler R package
data_path <- system.file("extdata", "SP_pheno_layout.xml", package = "SignalingProfiler")
RCy3::importVisualStyles(filename = data_path)
RCy3::setVisualStyle('SP_pheno_layout')

```

### Visualize network focused on experimentally validated phosphorylation events
This is the code used for **Figure 4A** of the manuscript.
```{r eval = FALSE}

RCy3::createNetworkFromIgraph(igraph=final_sp_visualization$igraph_network, name="PhosphoTest")

# Set in Cytoscape the SignalingProfiler style to focused on phosphorylation events 
# available in SignalingProfiler R package
data_path <- system.file("extdata", "SP_phospho_layout.xml", package = "SignalingProfiler")
RCy3::importVisualStyles(filename = data_path)
RCy3::setVisualStyle('SP_phospho_layout')

```

### Visualize functional circuits
This is the code used for **Figure 4B** and **Figure S9C** of the manuscript.

```{r eval = FALSE}

phenotypes <- c('APOPTOSIS', 'AUTOPHAGY', 'PROLIFERATION')
k_vector <- c(3, 4, 4)
start_nodes <- c('MTOR', 'AMPK')

for(i_pheno in 1:length(phenotypes)){
  
  circuit <- retrieve_functional_circuit(SP_object = final_sp_visualization, 
                                         start_nodes = start_nodes, 
                                         phenotype = phenotypes[i_pheno],
                                         k =  k_vector[i_pheno])
  
  RCy3::createNetworkFromIgraph(igraph=circuit, name= paste0(phenotypes[i_pheno], ' circuit'))
  
  # Apply on circuits the SignalingProfiler style
  data_path <- system.file("extdata", "SP_pheno_layout.xml", package = "SignalingProfiler")
  RCy3::importVisualStyles(filename = data_path)
  RCy3::setVisualStyle('SP_pheno_layout')
}
```




