---
title: "SignalingProfiler analysis with custom PKN"
author: "Veronica Venafra"
output:
  html_document:
    theme: flat
    highlight: zenburn
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#"
)
```

## Introduction

This workbook refers to the manuscript *SignalingProfiler 2.0: a network-based approach to bridge multi-omic data to phenotypic hallmarks*.

## Import SignalingProfiler 2.0 and other libraries

```{r}
library(SignalingProfiler)
library(tidyverse)
```

## Read input omic data

You can also embed plots, for example:

```{r pressure, echo=FALSE}
tr_df <- read_tsv('../input/transcriptomics.tsv')
prot_df <- read_tsv('../input/proteomics.tsv')
phospho_df <- read_tsv('../input/corrected_phosphoproteomics.tsv')
```

## Perform SignalingProfiler analysis with custom regulons

### Step 1) Infer the protein activity using custom regulons

We suggest to create all the regulons and then run the analysis, since the download of regulons requires some minutes. As an alternative, built-in regulons can be used (see SignalingProfiler tutorial).

#### TFEA analysis

```{r eval = FALSE}
tfea_regulons <- create_tfea_regulons(resources = c('SIGNOR', 'Dorothea'),
                                      organism = 'human')
write_tsv(tfea_regulons, '../custom_regulons/tfea_custom_regulons.tsv')
```

Run the analysis

```{r}
tfea_result <- run_footprint_based_analysis(omic_data = tr_df,
                                   analysis = 'tfea',
                                   organism = 'human',
                                   reg_minsize = 10,
                                   exp_sign = FALSE,
                                   integrated_regulons = FALSE,
                                   collectri = FALSE,
                                   hypergeom_corr = TRUE,
                                   correct_proteomics = TRUE,
                                   prot_df = prot_df,
                                   GO_annotation = TRUE,
                                   custom = TRUE,
                                   custom_path = '../custom_regulons/tfea_custom_regulons.tsv' )
```

#### KSEA analysis

```{r eval = FALSE}
ksea_regulons <- create_ksea_regulons(resources = c('SIGNOR', 'Omnipath'), 
                                      organism = 'human')
write_tsv(ksea_regulons, '../custom_regulons/ksea_custom_regulons.tsv')
```

Run the analysis

```{r}
ksea_result <- run_footprint_based_analysis(omic_data = phospho_df,
                                   analysis = 'ksea',
                                   organism = 'human',
                                   reg_minsize = 5,
                                   exp_sign = FALSE,
                                   integrated_regulons = FALSE,
                                   collectri = FALSE,
                                   hypergeom_corr = TRUE,
                                   correct_proteomics = TRUE,
                                   prot_df = prot_df,
                                   GO_annotation = TRUE,
                                   custom = TRUE,
                                   custom_path = '../custom_regulons/ksea_custom_regulons.tsv' )
```

#### PhosphoScore analysis

```{r eval = FALSE}
reg_phosphosites <- get_phosphoscore_info(resources = c('SIGNOR', 'PsP'), 
                      organism = 'human', 
                      psp_reg_site_path = '../../revisions/input/Regulatory_sites_2023-08-24',
                      only_activatory = TRUE)
write_tsv(reg_phosphosites, '../custom_regulons/act_phosphosites.tsv')
```

Run the PhosphoScore analysis

```{r}
phosphoscore_df <- phosphoscore_computation(phosphoproteomic_data = phospho_df,
                                            organism = 'human',
                                            activatory = TRUE ,
                                            GO_annotation= TRUE,
                                            custom = TRUE,
                                            custom_path = '../custom_regulons/act_phosphosites.tsv')
```

#### Final activity score computation

```{r}

combined_tfs <- combine_footprint_and_phosphoscore(footprint_output = tfea_result,
                                                   phosphoscore_df =  phosphoscore_df,
                                                   analysis =  'tfea')

combined_kin_phos <- combine_footprint_and_phosphoscore(footprint_output = ksea_result,
                                                        phosphoscore_df =  phosphoscore_df,
                                                        analysis =  'ksea')

toy_other <- phosphoscore_df %>%
  dplyr::filter(mf == 'other') %>%
  dplyr::rename(final_score = phosphoscore) %>%
  dplyr::mutate(method = 'PhosphoScore')

prot_activity_df <- dplyr::bind_rows(combined_tfs, 
                                    combined_kin_phos, 
                                    toy_other) %>%
  select(UNIPROT, gene_name, mf, final_score, method)

prot_activity_df
```

### Step 2) Context-specific network generation using custom PKN

#### Custom PKN generation
Generate a custom PKN choosing among 3 databases (4 if we consider Ser/Thr Kinome Atlas):

-   SIGNOR

-   PhosphoSitePlus: to use it you have to manually download 'Regulatory_sites.gz' and 'Kinase_Substrate_Dataset.gz' from [here](staticDownloads), after registration.

    -   You can also insert the information of [Ser/Thr Kinome Atlas work](https://www.nature.com/articles/s41586-022-05575-3).

-   OmniPath

You can also specify whether the interactions should be only direct with the direct argument.

```{r eval = FALSE}

create_PKN(database = c('SIGNOR', 'PsP'), #add also 'Omnipath' or 'Ser/Thr_Kinome_Atlas'
           direct = TRUE,
           organism = 'human', 
           psp_reg_site_path = '../../revisions/input/Regulatory_sites_2023-08-24',
           psp_kin_sub_path = '../../revisions/input/Kinase_Substrate_Dataset_2023-08-24', 
           file_path =  '../custom_regulons/custom_PKN.RDS') 
    
```

The function returns:

-   a list of an igraph object of the PKN;

-   a table of the entities included in the PKN;

-   a table of the interactions composing the PKN;

#### Naive network generation

```{r eval = FALSE}

# Extract the interactions object and apply SignalingProfiler commands
PKN_object <- readRDS('../custom_regulons/custom_PKN.RDS')
PKN_table <- PKN_object$interactions

PKN_expressed <- preprocess_PKN(omics_data = list(tr_df, prot_df, phospho_df),
                                PKN_table = PKN_table)

# divide proteins according to the molecular function
kin_phos_other <- prot_activity_df %>%
  dplyr::filter(mf %in% c('kin', 'phos', 'other'))
tfs <- prot_activity_df %>%
  dplyr::filter(mf == 'tf')

# create the naïve network
two_layers_toy <- two_layer_naive_network(starts_gn = c('AMPK', 'MTOR'),
                                          intermediate_gn = kin_phos_other$gene_name,
                                          targets_gn = tfs$gene_name,
                                          PKN_table = PKN_expressed, #or PKN_mouse
                                          max_length_1 = 3,
                                          max_length_2 = 4,
                                          rds_path = paste0('../custom_regulons/two_naive.rds'),
                                          sif_path = paste0('../custom_regulons/two_naive.sif'),
                                          connect_all = TRUE)

two_layers_toy
```
#### Optimization of the naive network over the protein activity

```{r eval = FALSE}
two_layers_toy <- readRDS('../custom_regulons/two_naive.rds')
receptor_list <- list('MTOR' = -1, 'AMPK' = 1)

carnival_input <- prepare_carnival_input(two_layers_toy,
                                         prot_activity_df,
                                         receptor_list,
                                         organism = 'human')

# Set ILP Solver options
solver = 'cplex'
carnival_options = default_CARNIVAL_options(solver)
carnival_options$timelimit <- 600

# FIRST RUN: RECEPTOR to KIN, PHOS, OTHERS
receptors_df <- carnival_input %>%
  dplyr::filter(mf == 'rec')

target1_df <- carnival_input%>%
  dplyr::filter(mf %in% c('kin', 'phos', 'other'))

naive_network <- readr::read_tsv(paste0('../custom_regulons/two_naive.sif'),
                                 col_names = c('source', 'interaction', 'target'))

output1 <- run_carnival_and_create_graph(source_df = receptors_df,
                                         target_df = target1_df,
                                         naive_network = unique(naive_network),
                                         proteins_df = carnival_input,
                                         organism = 'human',
                                         carnival_options = carnival_options,
                                         files = FALSE,
                                         with_atlas = FALSE)

# SECOND RUN: from KIN, PHOS, OTHERS to TFs
run1_output_nodes <- convert_output_nodes_in_next_input(output1)

source_df <- run1_output_nodes %>%
  dplyr::filter(mf %in% c('kin', 'phos', 'other'))
source_df$UNIPROT <- ''

target2_df <- carnival_input %>%
  dplyr::filter(mf == 'tf')

output2 <- run_carnival_and_create_graph(source_df = source_df,
                                         target_df = target2_df,
                                         naive_network = unique(naive_network),
                                         proteins_df = carnival_input,
                                         organism = 'human',
                                         carnival_options = carnival_options,
                                         files = FALSE,
                                         with_atlas = FALSE)

# UNION OF RUN1 and RUN2 graphs
union <- union_of_graphs(graph_1 = output1$igraph_network,
                         graph_2 = output2$igraph_network,
                         proteins_df = carnival_input,
                         files = TRUE,
                         path_sif = paste0('../custom_regulons/union_optimized.sif'),
                         path_rds = paste0('../custom_regulons/union_optimized.rds'))

# EXPANSION and MAPPING of PHOSPHOPROTEOMICS DATA ON EDGES
union_graph_rds <- readRDS(paste0('../custom_regulons/union_optimized.rds'))
union_output_final <- expand_and_map_edges(optimized_object = union_graph_rds,
                                           organism = 'human',
                                           phospho_df = phospho_df,
                                           files = TRUE,
                                           with_atlas = FALSE,
                                           path_sif = paste0('../custom_regulons/union_validated.sif'),
                                           path_rds = paste0('../custom_regulons/union_validated.rds'))

```

### Step 3) Phenotypic activity inference

```{r eval=FALSE}
pheno_table_distances <- phenoscore_network_preprocessing(proteomics = prot_df,
                                                          phospho = phospho_df)
write_rds(pheno_table_distances, '../custom_regulons/pheno_table_distances.rds')
```

```{r eval = FALSE}
pheno_table_distances <- readRDS('../custom_regulons/pheno_table_distances.rds')
sp_output <- readRDS('../custom_regulons/union_validated.rds')

toy_phenoscore_output<- phenoscore_computation(proteins_df = sp_output$nodes_df,
                                               desired_phenotypes = NULL,
                                               pheno_distances_table = pheno_table_distances,
                                               sp_graph = sp_output$igraph_network,
                                               # closeness of proteins to phenotypes
                                               path_length = 4,
                                               stat = 'mean',
                                               zscore_threshold = -1.96,
                                               # exclude random phenotypes
                                               n_random = 1000,
                                               pvalue_threshold = 0.05,
                                               # optimized network  specificity
                                               remove_cascade = TRUE,
                                               node_idx = TRUE,
                                               use_carnival_activity = TRUE,
                                               create_pheno_network = TRUE)

write_rds(toy_phenoscore_output, '../custom_regulons/phenoscore.rds')
```

```{r}

toy_phenoscore_output <- readRDS( '../custom_regulons/phenoscore.rds')

toy_phenoscore_output$table_phenotypes %>% 
  filter(EndPathways %in% c('fatty acid biosynthesis', 
                            'lipogenesis', 'autophagy', 
                            'apoptosis', 'cell death', 'adipogenesis', 
                            'glycogen synthesis', 
                            'protein synthesis', 
                            'glycolysis', 'proliferation')) %>% 
  arrange(desc(phenoscore))
```

